---
title: "Interactive Maps with leaflet in R DataCamp"
output: html_notebook
---

## Setting Up Interactive Web Maps
### Creating an Interactive Web Map
Similar to the packages in the tidyverse, the leaflet package makes use of the pipe operator (i.e., %>%) from the magrittr package to chain function calls together. This means we can pipe the result of one function into another without having to store the intermediate output in an object. For example, one way to find every car in the mtcars data set with a mpg >= 25 is to pipe the data through a series of functions.

```
mtcars  %>% 
    mutate(car = rownames(.))  %>% 
    select(car, mpg)  %>% 
    filter(mpg >= 25)  
```
    
To create a web map in R, you will chain together a series of function calls using the %>% operator. Our first function leaflet() will initialize the htmlwidget then we will add a map tile using the addTiles() function.

```{r}
# Load the leaflet library
library(leaflet)

# Create a leaflet map with default map tile using addTiles()
leaflet() %>%
    addTiles()
```

### Provider Tiles
In the previous exercise, addTiles() added the default OpenStreetMap (OSM) tile to your leaflet map. Map tiles weave multiple map images together. The map tiles presented adjust when a user zooms or pans the map enabling the interactive features you experimented with in exercise 2.

The leaflet package comes with more than 100 map tiles that you can use. These tiles are stored in a list called providers and can be added to your map using addProviderTiles() instead of addTiles().

```{r}

```

### 

## Plotting Points
### 

## Groups, Layers, and Extras
### 

## Plotting Polygons
### 