---
title: "Linear Algebra Datacamp"
output: html_notebook
---

### Creating Vectors in R
Vectors can be analyzed in a few different ways in R.

One way is to create the vector yourself, and there are a few different means by which you can do this.

The rep() command in R creates a vector that repeats the same element a prescribed number of times, while the seq() command creates a vector that follows a prescribed pattern.

A vector can be manually created by using the c() command, which stands for "concatenate".

```{r}
# Creating three 3's and four 4's, respectively
rep(3, 3)
rep(4, 4)

# Creating a vector with the first three even numbers and the first three odd numbers
seq(2, 6, by = 2)
seq(1, 5, by = 2)

# Re-creating the previous four vectors using the 'c' command
c(3, 3, 3)
c(4, 4, 4, 4)

c(2, 4, 6)
c(1, 3, 5)
```

### Creating Matrices in R
Matrices can be created and analyzed in a few different ways in R.

One way is to create the matrix yourself. There are a few different ways you can do this.

The matrix(a, nrow = b, ncol = c) command in R creates a matrix that repeats the element a in a matrix with b rows and c columns.

A matrix can be manually created by using the c() command as well.

```{r}
# Create a matrix of all 1's and all 2's that are 2 by 3 and 3 by 2, respectively
matrix(1, nrow = 2, ncol = 3)

print(matrix(2, nrow = 3, ncol = 2))

# Create a matrix B and changing the byrow designation.
B <- matrix(c(1, 2, 3, 2), nrow = 2, ncol = 2, byrow = FALSE)
B <- matrix(c(1, 2, 3, 2), nrow = 2, ncol = 2, byrow = TRUE)

# Add A to the previously-created matrix B
A + B
```
Matrix Multiplication as a Transformation
Matrices can be viewed as a way to transform collections of vectors into other vectors.

These transformations can take many forms, but the simplest ones in two dimensions are stretches or shrinkages (in either coordinate), reflections (e.g. about the x-axis, y-axis, origin, the line y = x), and rotations (clockwise, counter-clockwise).

Multiplication of a vector by a matrix is accomplished using the %*% command.

```{r}

```

